Sweat electrolytes and CFTR activity in the era of CFTR modulators

### 2023-02-08 Edith Zemanick/Sweat Test/sweat_test_analysis.Rmd
Update sweat_test_analysis.Rmd
tidyverse formatting

```{r setup, include=FALSE}
library(readxl)
library(arsenal)
library(tidyverse)
library(pROC)
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "/Users/timvigers/Documents/Work/CF/Edith Zemanick/Sweat Test")
source("~/GitHub/CF/get_genotype_class_severity.R")
```

```{r}
# Import exclusion data from Excel
diagnoses <- read_excel("./Data_Cleaned/AM reclassifications All Patients with CF or Non-CF.xlsx")
exclude <- diagnoses$PatientID[!is.na(diagnoses$Exclude)]
# Import data
df <- read.csv("./Data_Raw/Copy of Sweat Tests_AllFor202558_revised01Feb2021.csv", na.strings = "")
# Remove exclusions from Angela
df <- df[!(df$PatientID %in% exclude), ]
df$Diagnosis <- diagnoses$Diagnosis[match(df$PatientID, diagnoses$PatientID)]
# Add modulator info
cftr <- read.csv("./Data_Raw/Copy of Report CFTR Modulators.csv")
df$Modulator <- apply(df, 1, function(r) {
  mod <- cftr[cftr$Patient.ID == as.numeric(r["PatientID"]), ]
  if (any(duplicated(mod$Age.at.Start..years.))) {
    mod <- mod[-which(duplicated(mod$Age.at.Start..years.)), ]
  }
  mod <- mod[order(mod$Age.at.Start..years.), ]
  age <- as.numeric(r["Age.at.Test.in.Years"])
  as.character(cut(age, c(-Inf, mod$Age.at.Start..years., Inf),
    labels = c("None", mod$Modulator), right = F
  ))
})
# Numeric variables
num_vars = c("Amount","Na","K","Cl")
num_vars = c(paste0(num_vars,"_Left"),paste0(num_vars,"_Right"))
df[,num_vars] = suppressWarnings(lapply(df[,num_vars],as.numeric))
# Filter per Maxie's code
df <- df %>%
  mutate(Age.at.Test.in.Days = Age.at.Test.in.Weeks * 7) %>%
  filter(
    Age.at.Test.in.Years < 50 & Age.at.Test.in.Days >= 2,
    !(Diagnosis == "Non-CF" & PatientID %in%
      c(22, 23, 24, 29, 31, 474, 483, 484, 485, 505, 510, 785, 791, 792, 793)),
    Modulator == "None"
  ) %>%
  distinct(.) %>%
  filter(Sex != "", Demographics.Available == "TRUE")
# Genotype info
df$geno1 <- get_cf_genotype_class_severity(df$Genotypes1)
df$geno2 <- get_cf_genotype_class_severity(df$Genotypes2)
df$geno1[df$Diagnosis == "Non-CF"] <- "Not CF"
df$geno2[df$Diagnosis == "Non-CF"] <- "Not CF"
df$geno_risk <- paste(df$geno1, df$geno2, sep = "_")
df$geno_risk <- factor(df$geno_risk,
  levels = c(
    "Severe_Severe","Mild_Severe", "Mild_Unknown", "Severe_Mild", "Mild_Mild","Not CF_Not CF",

     "Severe_Unknown", "Unknown_Severe",
    "Unknown_Unknown"
  ),
  labels = c(
    "High","Low", "Low", "Low", "Low", "Not CF",  "Unknown",
    "Unknown", "Unknown"
  )
)
# Which arm?
df$Arm <- apply(df[, c("Cl", "Cl_Left", "Cl_Right", "Na", "Na_Left", "Na_Right", "K", "K_Left", "K_Right")], 1, function(r) {
  k <- sum(!is.na(r))
  if (k == 6) {
    return("Both")
  } else if (k == 3) {
    return("Arm Not Noted")
  } else {
    return(NA)
  }
})

# Difference between arms
df$`Cl Diff. Between Left and Right Arm` = as.numeric(df$Cl_Left) - as.numeric(df$Cl_Right)
df$`Na Diff. Between Left and Right Arm` = as.numeric(df$Na_Left) - as.numeric(df$Na_Right)
# Merge left and right values
df$Amount <- apply(df[, c("Amount", "Amount_Left", "Amount_Right")], 1, function(r) {
  amt <- suppressWarnings(as.numeric(as.character(r)))
  mean(amt, na.rm = T)
})
df$Cl <- apply(df[, c("Cl", "Cl_Left", "Cl_Right")], 1, function(r) {
  cl <- suppressWarnings(as.numeric(as.character(r)))
  mean(cl, na.rm = T)
})
df$Na <- apply(df[, c("Na", "Na_Left", "Na_Right")], 1, function(r) {
  na <- suppressWarnings(as.numeric(as.character(r)))
  mean(na, na.rm = T)
})
df$K <- apply(df[, c("K", "K_Left", "K_Right")], 1, function(r) {
  k <- suppressWarnings(as.numeric(as.character(r)))
  mean(k, na.rm = T)
})
# Sodium to chloride ratio
df$`Na/Cl Ratio` = df$Na /df$Cl
# Sort
df <- df %>% arrange(PatientID, Age.at.Test.in.Weeks)
# Table options (most variables are not normally distributed)
mycontrols  <- tableby.control(numeric.test="kwt",numeric.stats=c("Nmiss", "median", "q1q3"))
labels = list(Age.at.Test.in.Years = "Age (Years)", geno_risk = "Genotype Severity")
```

@@

```{r results='asis'}
t1_df <- df %>%
  group_by(PatientID) %>%
  filter(row_number() == 1) # First measure for everyone
t1 <- tableby(Diagnosis ~ Sex + Age.at.Test.in.Years + fe(geno_risk) + Amount + Cl + Na + K + `Na/Cl Ratio`,
  data = t1_df,control = mycontrols,simulate.p.value=T
)
summary(t1,pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(Method ~ Sex + Age.at.Test.in.Years + Diagnosis + geno_risk + Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df,control = mycontrols,
)
summary(t1,pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(Diagnosis ~ Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df,control = mycontrols,
)
summary(t1,pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(Diagnosis ~ Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df[t1_df$Method == "Gibson-Cooke",],control = mycontrols,
)
summary(t1,pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(Diagnosis ~ Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df[t1_df$Method == "Macroduct",],control = mycontrols,
)
summary(t1,pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(geno_risk ~ Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df,control = mycontrols,
)
summary(t1,pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(geno_risk ~ Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df[t1_df$Diagnosis == "Cystic Fibrosis",],control = mycontrols
)
summary(t1,pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(geno_risk ~ Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df[t1_df$Diagnosis == "CFTR-related disorder",],control = mycontrols
)
summary(t1,pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(geno_risk ~ Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df[t1_df$Diagnosis == "CRMS",],control = mycontrols
)
summary(t1,pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby( ~ Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df[t1_df$Diagnosis == "Non-CF",],control = mycontrols
)
summary(t1,pfootnote = T,
  labelTranslations = labels
)
```

## Plots

@@

@@

```{r results='asis'}
# Get first and last values for everyone with more than 1
multiples = df %>% group_by(PatientID) %>% 
  filter(n() > 1) %>% # Multiple values only
  filter(row_number()==1 | row_number()==n()) %>% # First and last
  mutate(time = row_number())
multiples$time = factor(multiples$time,levels = 1:2,labels = c("First","Last"))
# Paired testing options
mycontrols  <- tableby.control(numeric.test="signed.rank",cat.test = "mcnemar",
                               numeric.stats=c("Nmiss", "median", "q1q3"))
```

@@

```{r results='asis'}
p <- paired(time ~ Diagnosis +Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,signed.rank.exact = F,
            data = multiples, id = PatientID,control = mycontrols)
summary(p,pfootnote = T,labelTranslations = labels)
```

@@

```{r results='asis'}
p <- paired(time ~ Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,signed.rank.exact = F,strata = Diagnosis,
            data = multiples %>% filter(Diagnosis %in% c("Cystic Fibrosis","Non-CF")), 
            id = PatientID,control = mycontrols)
summary(p,pfootnote = T,labelTranslations = labels)
```

@@

```{r results='asis'}
p <- paired(time ~ Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,signed.rank.exact = F,strata = geno_risk,
            data = multiples, 
            id = PatientID,control = mycontrols)
summary(p,pfootnote = T,labelTranslations = labels)
```

@@

```{r results='asis'}
p <- paired(time ~ Cl + Na + K + `Na/Cl Ratio`,signed.rank.exact = F,strata = geno_risk,
            data = multiples %>% filter(Diagnosis == "Cystic Fibrosis"), 
            id = PatientID,control = mycontrols)
summary(p,pfootnote = T,labelTranslations = labels)
```

@@

```{r results='asis'}
p <- paired(time ~ Cl + Na + K + `Na/Cl Ratio`,signed.rank.exact = F,strata = geno_risk,
            data = multiples %>% filter(Diagnosis == "Non-CF"), 
            id = PatientID,control = mycontrols)
summary(p,pfootnote = T,labelTranslations = labels)
```

@@

```{r}
# Three groups
t1_df$Group = cut(t1_df$Cl,breaks = c(0,30,60,Inf),right = F,labels = c("Low","Medium","High"))
# Low vs. medium or high
t1_df$Low = t1_df$Group
levels(t1_df$Low) = c("Low","Medium or High","Medium or High")
t1_df$Low = relevel(t1_df$Low,ref = "Medium or High")
# High vs. medium or low
t1_df$High = t1_df$Group
levels(t1_df$High) = c("Low or Medium","Low or Medium","High")
t1_df$High = relevel(t1_df$High,ref = "Low or Medium")
```

@@

```{r}
roc_low <- roc(response = t1_df$Low,predictor = t1_df$Na,quiet = T)
coords_low = round(coords(roc_low,"best",best.method="youden"),3)
roc_high <- roc(response = t1_df$High,predictor = t1_df$Na,quiet = T)
coords_high = round(coords(roc_high,"best",best.method="youden"),3)

plot(roc_low,print.auc = TRUE, col = "blue") 
plot(roc_high, print.auc = TRUE, col = "green", print.auc.y = .4, add = TRUE)
```

@@

```{r warning=FALSE}
ggplot(t1_df[!is.na(t1_df$Group),],aes(Na,fill = Group)) + 
  geom_histogram(alpha=0.3, position="identity",binwidth = 1) + 
  geom_vline(xintercept = coords_low$threshold,color = "red") +
  geom_vline(xintercept = coords_high$threshold,color = "red") +
  ylab("Count") + theme_bw()
```

>>>>

```{r setup, include=FALSE}
library(readxl)
library(arsenal)
library(tidyverse)
library(pROC)
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
if (Sys.info()["sysname"] == "Windows") {
  home_dir <- "C:/Users/timvigers/OneDrive - The University of Colorado Denver/Vigers/CF/Edith Zemanick/Sweat Test"
  source("C:/Users/timvigers/Documents/GitHub/CF/get_genotype_class_severity.R")
} else if (Sys.info()["sysname"] == "Linux") {
  home_dir <- "~/UCD/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Edith Zemanick/Sweat Test"
  source("~/GitHub/CF/get_genotype_class_severity.R")
} else if (Sys.info()["sysname"] == "Darwin") {
  home_dir <- "/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Edith Zemanick/Sweat Test"
  source("~/GitHub/CF/genotype_class_severity.R")
}
knitr::opts_knit$set(root.dir = home_dir)
rm(home_dir)
```

```{r}
# Import exclusion data from Excel
diagnoses <- read_excel("./Data_Cleaned/AM reclassifications All Patients with CF or Non-CF.xlsx")
exclude <- diagnoses$PatientID[!is.na(diagnoses$Exclude)]
# Import data
df <- read.csv("./Data_Raw/Copy of Sweat Tests_AllFor202558_revised01Feb2021.csv", na.strings = "")
# Remove exclusions from Angela
df <- df[!(df$PatientID %in% exclude), ]
df$Diagnosis <- diagnoses$Diagnosis[match(df$PatientID, diagnoses$PatientID)]
# Add modulator info
cftr <- read.csv("./Data_Raw/Copy of Report CFTR Modulators.csv")
df$Modulator <- apply(df, 1, function(r) {
  mod <- cftr[cftr$Patient.ID == as.numeric(r["PatientID"]), ]
  if (any(duplicated(mod$Age.at.Start..years.))) {
    mod <- mod[-which(duplicated(mod$Age.at.Start..years.)), ]
  }
  mod <- mod[order(mod$Age.at.Start..years.), ]
  age <- as.numeric(r["Age.at.Test.in.Years"])
  as.character(cut(age, c(-Inf, mod$Age.at.Start..years., Inf),
    labels = c("None", mod$Modulator), right = F
  ))
})
# Numeric variables
num_vars <- c("Amount", "Na", "K", "Cl")
num_vars <- c(paste0(num_vars, "_Left"), paste0(num_vars, "_Right"))
df[, num_vars] <- suppressWarnings(lapply(df[, num_vars], as.numeric))
# Filter per Maxie's code
df <- df %>%
  mutate(Age.at.Test.in.Days = Age.at.Test.in.Weeks * 7) %>%
  filter(
    Age.at.Test.in.Years < 50 & Age.at.Test.in.Days >= 2,
    !(Diagnosis == "Non-CF" & PatientID %in%
      c(22, 23, 24, 29, 31, 474, 483, 484, 485, 505, 510, 785, 791, 792, 793)),
    Modulator == "None"
  ) %>%
  distinct(.) %>%
  filter(Sex != "", Demographics.Available == "TRUE")
# Genotype info
## By severity
df$geno1 <- cf_genotype_class_severity(df$Genotypes1)
df$geno2 <- cf_genotype_class_severity(df$Genotypes2)
df$geno1[df$Diagnosis == "Non-CF"] <- "Not CF"
df$geno2[df$Diagnosis == "Non-CF"] <- "Not CF"
df$geno_risk <- paste(df$geno1, df$geno2, sep = "_")
df$geno_risk <- factor(df$geno_risk,
  levels = c(
    "Severe_Severe", "Mild_Severe", "Mild_Unknown", "Severe_Mild", "Mild_Mild",
    "Not CF_Not CF", "Severe_Unknown", "Unknown_Severe", "Unknown_Unknown"
  ),
  labels = c(
    "High", "Low", "Low", "Low", "Low", "Not CF", "Unknown",
    "Unknown", "Unknown"
  )
)
## By function
df$geno1_class <- cf_genotype_class_severity(df$Genotypes1, out = "class")
df$geno1_class[grep("508", df$Genotypes1)] <- "F508del"
df$geno1_class <- replace(df$geno1_class, df$geno1_class %in% c("I", "II", "III"), "MF")
df$geno1_class <- replace(df$geno1_class, df$geno1_class %in% c("IV", "V"), "RF")

df$geno2_class <- cf_genotype_class_severity(df$Genotypes2, out = "class")
df$geno2_class[grep("508", df$Genotypes2)] <- "F508del"
df$geno2_class <- replace(df$geno2_class, df$geno2_class %in% c("I", "II", "III"), "MF")
df$geno2_class <- replace(df$geno2_class, df$geno2_class %in% c("IV", "V"), "RF")

df$geno_class <- factor(paste0(
  pmin(df$geno1_class, df$geno2_class), "/",
  pmax(df$geno1_class, df$geno2_class)
))
# Which arm?
df$Arm <- apply(df[, c("Cl", "Cl_Left", "Cl_Right", "Na", "Na_Left", "Na_Right", "K", "K_Left", "K_Right")], 1, function(r) {
  k <- sum(!is.na(r))
  if (k == 6) {
    return("Both")
  } else if (k == 3) {
    return("Arm Not Noted")
  } else {
    return(NA)
  }
})

# Difference between arms
df$`Cl Diff. Between Left and Right Arm` <- as.numeric(df$Cl_Left) - as.numeric(df$Cl_Right)
df$`Na Diff. Between Left and Right Arm` <- as.numeric(df$Na_Left) - as.numeric(df$Na_Right)
# Merge left and right values
df$Amount <- apply(df[, c("Amount", "Amount_Left", "Amount_Right")], 1, function(r) {
  amt <- suppressWarnings(as.numeric(as.character(r)))
  mean(amt, na.rm = T)
})
df$Cl <- apply(df[, c("Cl", "Cl_Left", "Cl_Right")], 1, function(r) {
  cl <- suppressWarnings(as.numeric(as.character(r)))
  mean(cl, na.rm = T)
})
df$Na <- apply(df[, c("Na", "Na_Left", "Na_Right")], 1, function(r) {
  na <- suppressWarnings(as.numeric(as.character(r)))
  mean(na, na.rm = T)
})
df$K <- apply(df[, c("K", "K_Left", "K_Right")], 1, function(r) {
  k <- suppressWarnings(as.numeric(as.character(r)))
  mean(k, na.rm = T)
})
# Sodium to chloride ratio
df$`Na/Cl Ratio` <- df$Na / df$Cl
# Sort
df <- df %>% arrange(PatientID, Age.at.Test.in.Weeks)
# Table options (most variables are not normally distributed)
mycontrols <- tableby.control(numeric.test = "kwt", numeric.stats = c("Nmiss", "median", "q1q3"))
labels <- list(
  Age.at.Test.in.Years = "Age (Years)",
  geno_risk = "Genotype Severity", geno_class = "Specific Genotype"
)
```

@@

```{r results='asis'}
t1_df <- df %>%
  group_by(PatientID) %>%
  filter(row_number() == 1) # First measure for everyone
t1 <- tableby(
  Diagnosis ~ Sex + Age.at.Test.in.Years + fe(geno_risk) + fe(geno_class) +
    Amount + Cl + Na + K + `Na/Cl Ratio`,
  data = t1_df, control = mycontrols, simulate.p.value = T
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(
  Method ~ Sex + Age.at.Test.in.Years + Diagnosis + fe(geno_risk) + fe(geno_class) +
    Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` +
    `Na Diff. Between Left and Right Arm`,
  data = t1_df, control = mycontrols, simulate.p.value = T
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(
  Diagnosis ~ Amount + Cl + Na + K + `Na/Cl Ratio` +
    `Cl Diff. Between Left and Right Arm` +
    `Na Diff. Between Left and Right Arm`,
  data = t1_df, control = mycontrols,
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(
  Diagnosis ~ Amount + Cl + Na + K + `Na/Cl Ratio` +
    `Cl Diff. Between Left and Right Arm` +
    `Na Diff. Between Left and Right Arm`,
  data = t1_df[t1_df$Method == "Gibson-Cooke", ], control = mycontrols,
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(
  Diagnosis ~ Amount + Cl + Na + K + `Na/Cl Ratio` +
    `Cl Diff. Between Left and Right Arm` +
    `Na Diff. Between Left and Right Arm`,
  data = t1_df[t1_df$Method == "Macroduct", ], control = mycontrols,
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(
  geno_risk ~ Amount + Cl + Na + K + `Na/Cl Ratio` +
    `Cl Diff. Between Left and Right Arm` +
    `Na Diff. Between Left and Right Arm`,
  data = t1_df, control = mycontrols,
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(geno_risk ~ Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df[t1_df$Diagnosis == "Cystic Fibrosis", ], control = mycontrols
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(geno_risk ~ Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df[t1_df$Diagnosis == "CFTR-related disorder", ], control = mycontrols
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(geno_risk ~ Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df[t1_df$Diagnosis == "CRMS", ], control = mycontrols
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(~ Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df[t1_df$Diagnosis == "Non-CF", ], control = mycontrols
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

### Plots

@@

## By specific genotype

```{r results='asis'}
t1 <- tableby(
  geno_class ~ Amount + Cl + Na + K + `Na/Cl Ratio` +
    `Cl Diff. Between Left and Right Arm` +
    `Na Diff. Between Left and Right Arm`,
  data = t1_df, control = mycontrols,
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

## By specific genotype and diagnosis

```{r results='asis'}
t1 <- tableby(geno_class ~ Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df[t1_df$Diagnosis == "Cystic Fibrosis", ], control = mycontrols
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

```{r results='asis'}
t1 <- tableby(geno_class ~ Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df[t1_df$Diagnosis == "CFTR-related disorder", ], control = mycontrols
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

```{r results='asis'}
t1 <- tableby(geno_class ~ Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df[t1_df$Diagnosis == "CRMS", ], control = mycontrols
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

```{r results='asis'}
t1 <- tableby(~ Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df[t1_df$Diagnosis == "Non-CF", ], control = mycontrols
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

### Plots

```{r warning=FALSE}
ggplot(df, aes(x = geno_class, y = Na)) +
  geom_boxplot() +
  theme_bw() +
  xlab("Specific Genotype") +
  ggtitle("Sweat Sodium")

ggplot(df, aes(x = geno_class, y = K)) +
  geom_boxplot() +
  theme_bw() +
  ggtitle("Sweat Potassium")

ggplot(df, aes(x = geno_class, y = Cl)) +
  geom_boxplot() +
  theme_bw() +
  xlab("Specific Genotype") +
  ggtitle("Sweat Chloride")

ggplot(df, aes(x = geno_class, y = `Na/Cl Ratio`)) +
  geom_boxplot() +
  theme_bw() +
  xlab("Specific Genotype") +
  ggtitle("Na/Cl Ratio")
```

@@

```{r results='asis'}
# Get first and last values for everyone with more than 1
multiples <- df %>%
  group_by(PatientID) %>%
  filter(n() > 1) %>% # Multiple values only
  filter(row_number() == 1 | row_number() == n()) %>% # First and last
  mutate(time = row_number())
multiples$time <- factor(multiples$time, levels = 1:2, labels = c("First", "Last"))
# Paired testing options
mycontrols <- tableby.control(
  numeric.test = "signed.rank", cat.test = "mcnemar",
  numeric.stats = c("Nmiss", "median", "q1q3")
)
```

@@

```{r results='asis'}
p <- paired(
  time ~ Diagnosis + Cl + Na + K + `Na/Cl Ratio` +
    `Cl Diff. Between Left and Right Arm` +
    `Na Diff. Between Left and Right Arm`,
  signed.rank.exact = F, data = multiples,
  id = PatientID, control = mycontrols
)
summary(p, pfootnote = T, labelTranslations = labels)
```

@@

```{r results='asis'}
p <- paired(time ~ Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  signed.rank.exact = F, strata = Diagnosis,
  data = multiples %>% filter(Diagnosis %in% c("Cystic Fibrosis", "Non-CF")),
  id = PatientID, control = mycontrols
)
summary(p, pfootnote = T, labelTranslations = labels)
```

@@

```{r results='asis'}
p <- paired(time ~ Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  signed.rank.exact = F, strata = geno_risk,
  data = multiples,
  id = PatientID, control = mycontrols
)
summary(p, pfootnote = T, labelTranslations = labels)
```

@@

```{r results='asis'}
p <- paired(time ~ Cl + Na + K + `Na/Cl Ratio`,
  signed.rank.exact = F, strata = geno_risk,
  data = multiples %>% filter(Diagnosis == "Cystic Fibrosis"),
  id = PatientID, control = mycontrols
)
summary(p, pfootnote = T, labelTranslations = labels)
```

@@

```{r results='asis'}
p <- paired(time ~ Cl + Na + K + `Na/Cl Ratio`,
  signed.rank.exact = F, strata = geno_risk,
  data = multiples %>% filter(Diagnosis == "Non-CF"),
  id = PatientID, control = mycontrols
)
summary(p, pfootnote = T, labelTranslations = labels)
```

## By specific genotype

```{r results='asis'}
p <- paired(
  time ~ Cl + Na + K + `Na/Cl Ratio` +
    `Cl Diff. Between Left and Right Arm` +
    `Na Diff. Between Left and Right Arm`,
  signed.rank.exact = F, strata = geno_class, numeric.test = "notest",
  data = multiples, id = PatientID, control = mycontrols
)
summary(p, pfootnote = T, labelTranslations = labels)
```

## By genotype class and diagnosis

```{r results='asis'}
p <- paired(time ~ Cl + Na + K + `Na/Cl Ratio`,
  signed.rank.exact = F, strata = geno_class,
  data = multiples %>% filter(Diagnosis == "Cystic Fibrosis"),
  id = PatientID, control = mycontrols
)
summary(p, pfootnote = T, labelTranslations = labels)
```

```{r results='asis'}
p <- paired(time ~ Cl + Na + K + `Na/Cl Ratio`,
  signed.rank.exact = F, strata = geno_class,
  data = multiples %>% filter(Diagnosis == "Non-CF"),
  id = PatientID, control = mycontrols
)
summary(p, pfootnote = T, labelTranslations = labels)
```

@@

```{r}
# Three groups
t1_df$Group <- cut(t1_df$Cl, breaks = c(0, 30, 60, Inf), right = F, labels = c("Low", "Medium", "High"))
# Low vs. medium or high
t1_df$Low <- t1_df$Group
levels(t1_df$Low) <- c("Low", "Medium or High", "Medium or High")
t1_df$Low <- relevel(t1_df$Low, ref = "Medium or High")
# High vs. medium or low
t1_df$High <- t1_df$Group
levels(t1_df$High) <- c("Low or Medium", "Low or Medium", "High")
t1_df$High <- relevel(t1_df$High, ref = "Low or Medium")
```

@@

```{r}
roc_low <- roc(response = t1_df$Low, predictor = t1_df$Na, quiet = T)
coords_low <- round(coords(roc_low, "best", best.method = "youden"), 3)
roc_high <- roc(response = t1_df$High, predictor = t1_df$Na, quiet = T)
coords_high <- round(coords(roc_high, "best", best.method = "youden"), 3)

plot(roc_low, print.auc = TRUE, col = "blue")
plot(roc_high, print.auc = TRUE, col = "green", print.auc.y = .4, add = TRUE)
```

@@

```{r warning=FALSE}
ggplot(t1_df[!is.na(t1_df$Group), ], aes(Na, fill = Group)) +
  geom_histogram(alpha = 0.3, position = "identity", binwidth = 1) +
  geom_vline(xintercept = coords_low$threshold, color = "red") +
  geom_vline(xintercept = coords_high$threshold, color = "red") +
  ylab("Count") +
  theme_bw()
```
